---
title: "W1-Self-Assessment (Part II)"
subtitle: "Jaidev Deshpande\nRoll No: 21F1003751"
output:
  pdf_document: default
  html_notebook: default
---

## Week 1: Self-Assessment (Part II)

---

1. Two sets of values:
$$
x_0 = 10, a = 4^5, m = 2^{20} - 1, c = 2^5 - 1 \\
x_0 = 10, a = 8^5, m = 2^{41} - 1, c = 2^8 - 1
$$

  a. Write an R code to generate 1000 pseudorandom numbers using the ‘Mixed Con-gruential Method’ for the first set of values. Also, plot the histogram and trace plot for the above 1000 generated pseudorandom numbers.
  
```{r}
mixed_congruential <- function(
    seed, n_samples, a = 7^5, m = (2 ^ 31) - 1, c=(2 ^ 10) - 1) {
  x <- rep(NA, times=n_samples)
  x[1] <- seed
  hasRepeat <- FALSE
  for (i in 2:n_samples) {
    val <- (x[i - 1] * a + c) %% m
    # If val is already present, the values will repeat
    if (val %in% x[1:i]) {
      hasRepeat <- TRUE
      break
    }
    x[i] <- val
  }
  if (hasRepeat) {  # loop broke because of repeating values, so fill the rest
    reps <- x[1:(i - 1)]  # the original sequence to be repeated
    x[i:n_samples] <- rep(reps, length.out=(n_samples - i + 1))
  }
  return(x / m)
}

# First set of values
x <- mixed_congruential(10, 1000, a=4^5, m=(2 ^ 20)-1, c=(2^5)-1)
par(mfrow=c(1, 2))
hist(x)
plot.ts(x)
```


  b. Second set of values

```{r}
x <- mixed_congruential(10, 1000, a=8^5, m=(2 ^ 41)-1, c=(2^8)-1)
par(mfrow=c(1, 2))
hist(x)
plot.ts(x)
```

  c. Comments & inference
      Hello World.
      
---

2.(a) Generate a sample from Binomial distribution with parameters
n = 5 and p = 0.3 by using the “Inverse Transform Method”.

```{r}
uniform_sample <- runif(1)

# params for Binomail distribution
n = 5
p = 0.3
cdf <- unique(pbinom(-1:5, n, p))
binomial_sample <- cut(uniform_sample, cdf, labels=FALSE)
print(binomial_sample)
```

2.(b) Generate 200 samples

```{r}
uniform_samples <- runif(200)
binomial_samples <- cut(uniform_samples, cdf, labels=FALSE) - 1
print(mean(binomial_samples))  # empirical
print(n * p)  # theoretical
```


---

3. To get samples from a binomial distribution with params (n, p) using the AR sampler.
# Re-evaluate the following:
3.(i)
  a. Bernoulli - **cannot** be used since the support of this distribution is _smaller_ than that of the binomial distribution
  b. Geometric - **cannot** be used since the proposal support is a _subset_ of the target support (number of trials until the first success cannot be zero.)
  c. Negative binomial - **can** be used since it has the same support.
  d. Hypergeometric - **cannot** be used since it has a countably finite support.

3.(ii)  Expression for c (upper bound) when the proposal is Poisson.