---
title: "W1-Self-Assessment (Part I)"
subtitle: "Jaidev Deshpande\nRoll No: 21F1003751"
output:
  pdf_document: default
  html_notebook: default
---

## Week 1: Self-Assessment (Part I)

---

1.(i) Using R, create a variable named n and assign a random value to it.

```{r}
n <- runif(1, min=-1, max=1)
print(n)
```


1.(ii) Write an R code to check the sign of the above assigned value to n. Print the following as output:

  * “Number is positive” if the assigned value is positive.
  * “Number is negative” if the assigned value is negative.
  * “Number is zero” if the assigned value is zero.

```{r}
if (n > 0) {
  print('Number is positive')
} else if (n < 0) {
  print('Number is negative')
} else { print('Number is zero') }
```


1.(iii) Write an R code to count the number of positive and negative entries in the vector `c(-1, 2, 12, -4, 8, 9)`.

```{r}
x <- c(-1, 2, 12, -4, 8, 9)
sprintf("No. of positives: %d", sum(x > 0))
sprintf("No. of negatives: %d", sum(x < 0))
```


---

2.(i) Write a function using R to find the area of a circle which takes in parameter “radius”.

```{r}
circle_area <- function(radius) {
  return(pi * radius * radius)
}
circle_area(1)
```


2.(ii) Using the above function, find the area of the circle with radius 4.

```{r}
circle_area(4)
```

---

3.(i) Write a function using R to get the first `n` Fibonacci numbers, which takes in parameter `n`.

```{r}
CACHE <- setNames(data.frame(c(0, 1)), "nFib")  # Use a cache to look up previous Fibonacci numbers
fibonacci <- function(n) {
  if (is.na(CACHE$nFib[n + 1])) {
    nth_val <- fibonacci(n - 1) + fibonacci(n - 2)
    CACHE[n + 1, "nFib"] <<- nth_val
  }
  return(CACHE[n + 1, "nFib"])
}
```

3.(ii) Using the above function, print the first 10 Fibonacci numbers.
```{r}
foo <- fibonacci(10)
print(CACHE$nFib[2:11])
```

---

4.(i) Using R, create an empty matrix A of size 3 $\times$ 3

```{r}
A <- matrix(nrow=3, ncol=3)
```

4.(ii) Fill the entries of A row-wise with the vector c(1, 2, 3, 4, 5, 6, 7, 8, 9) and print matrix A.

```{r}
A <- matrix(1:9, nrow=3, ncol=3, byrow=TRUE)
print(A)
```

4.(iii) Concatenate the vector c(10, 11, 12) as a new row to the existing matrix A and
store it in a new matrix B.

```{r}
B <- rbind(A, c(10, 11, 12))
print(B)
```

4.(iv) Construct a 4 $\times$ 1 matrix which takes the input as vector c(1, 2, 3, 4) and store it
as C. Concatenate C as a new column to matrix B and store it in matrix X.

```{r}
C <- matrix(1:4, nrow=4, ncol=1)
X <- cbind(B, C)
print(X)
```

4.(v) Print the first and the third column of matrix X.

```{r}
print(X[, 1])
print(X[, 3])
```

4.(vi) Create a 4 × 4 matrix with input vector as c(1 : 16) and store it as Y.

```{r}
Y <- matrix(1:16, 4, 4, byrow=TRUE)  # Assuming from the earlier question that this is row-wise
print(Y)
```

4.(vi)(a) Transpose of Y

```{r}
Y_transpose <- t(Y)
print(Y_transpose)
```

4.(vi)(b) Elementwise sum of X and Y

```{r}
Z <- X + Y
print(Z)
```

4.(vi)(c) Elementwise product of X and Y

```{r}
P <- X * Y
print(P)
```

4.(vi)(d) Matrix multiplication of X and Y

```{r}
I <- X %*% Y
print(I)
```

---