---
title: 'Week 6: Graded Assignment'
author: "Jaidev Deshpande"
date: "2023-07-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Estimate $\theta = \int_{0}^{2}sin(x)f(x)dx$ where $f(x) \propto \frac{1}{1 + x^2}$ and $g(x) = \frac{x}{2} \forall x \in [0, 2]$.

    Following weighted importance sampling, we can set the following functions:
    
    $$
    \tilde{f(x)} = \frac{1}{1 + x^2}\\
    \tilde{g(x)} = x\\
    h(x) = sin(x)
    $$
    
    1.(a) Weights for drawing from $g$ are expressed as:
  
    $$
    w(x) = \frac{\tilde{f(x)}}{\tilde{g(x)}} = \frac{1}{x(1 + x^2)}
    $$
    
    1.(b) Function to generate weights
    ```{r}
    weights <- function(N) {
      xx <- runif(N, min=0, max=2)
      return(1 / (xx * (1 + xx ^ 2)))
    }
    ```
  
    1.(c) Estimated value of $\theta$ for $N = 10000$:

    ```{r}
    N <- 10000
    
    weighted.IS <- function(N) {
      hx <- sin(runif(N, min=0, max=2))
      weights <- weights(N)
      theta <- sum(hx * weights) / sum(weights)
      return(theta)
    }
    ```

---

2. Linear Regression on the Iris dataset:

    2.(a) Scatterplot of sepal length vs petal length.

    ```{r}
    plot(iris$Sepal.Length, iris$Petal.Length, xlab="Sepal Length", ylab="Petal Length")
    ```

    Simple linear regression is somewhat (not perfectly) suitable for this problem,
    as there is an _approximately_ linear relationship between sepal length and petal length.
    It will not be an optimal model in the squared error sense.

    2.(b) MLE for the slope and the intercept
    
    ```{r}
    
    X <- cbind(1, iris$Sepal.Length)
    y <- iris$Petal.Length
    
    beta.hat <- solve(t(X) %*% X) %*% t(X) %*% y
    
    intercept <- beta.hat[1,]
    slope <- beta.hat[2,]
    print(c(slope, intercept))
    ```
    
    2.(c) MLE estimate of $\sigma^2$
    
    ```{r}
    var.est <- t(y - X %*% beta.hat) %*% (y - X %*% beta.hat) / length(y)
    print(var.est[,1])
    ```
    
    2.(d) Best line fit:
    
    
    ```{r}
    plot(iris$Sepal.Length, iris$Petal.Length, xlab="Sepal Length", ylab="Petal Length")
    lines(iris$Sepal.Length, iris$Sepal.Length * slope + intercept, col="blue")
    ```