---
title: 'Week 4: Graded Assignment'
author: |
  | Jaidev Deshpande
  | Roll No: 21F1003751

output:
  pdf_document: default
---

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Generate samples from the distribution $f(x) = 30(x^2 -2x^3 + x^4); x \in [0, 1]$

    (i). **What will be the set $C$ for the above $f(x)$?**
    
    The set $C$ for any distribution $f(x)$ is defined as follows,
    
    $$
    C = \left\{(u, v): 0 \leq u \leq \sqrt{f\left(\frac{v}{u}\right)}\right\}
    $$
    
    Simplifying the given distribution:
    $$f(x) = 30(x^2 -2x^3 + x^4)$$
    $$\therefore f(x) = 30x^2(1 - 2x + x^2)$$
    $$\therefore f(x) = 30x^2(1 - x)^2$$
    $$\therefore \sqrt{f(x)} = \sqrt{30}x(1-x)$$
    
    Thus, for this distribution, the set $C$ is,
    
    $$
    C = \left\{(u, v): 0 \leq u \leq \sqrt{30}\frac{v}{u}\left(1 - \frac{v}{u}\right)\right\}
    $$

    (ii). **The bounds of the rectangle $[a, b] \times [c, d]$**
    
    $a$ is, as usual, $0$.
    
    $$a:=0$$
    
    $b$ is obtained as follows:
    
    $$ b:= \max_{x \in [0, 1]} \sqrt{f(x)} $$
    $$ \therefore b= \max_{x \in [0, 1]} x (1 - x) \sqrt{30}$$
    $$ \therefore b= \frac{\sqrt{30}}{4} \approx 1.369 $$
    
    $c$ is obtained as:
    
    $$ c := \inf_{x\leq0} x \sqrt{f(x)} $$
    $$ \therefore c = 0 $$
    
    $d$ is obtained as:
    
    $$ d := \sup_{x\geq0} x \sqrt{f(x)} $$
    $$ \therefore d = 0.811 $$
    
    Thus, the rectangle $[a, b] \times [c, d]$ is $[0, 1.369] \times [0, 0.811]$
    
    (iv). **Generate 10,000 samples from $f(x)$.**
    
    ```{r}
    sqrt.fx <- function(x) {
      f <- (x ^ 4) - 2 * (x ^ 3) + (x ^ 2)
      return(sqrt(30 * f))
    }

    rou.sample <- function(a, b, c, d) {
      # Return a sample from the rectangle [a, b, c, d], and number of tries
      n.try <- 0
      while (TRUE) {
        n.try <- n.try + 1
        u <- runif(1, min=a, max=b)
        v <- runif(1, min=c, max=d)
        ratio <- v / u
        if (u <= sqrt.fx(ratio) && ratio < 1) {
          break
        }
      }
      return(c(ratio, n.try))
    }
    
    N <- 10000
    samples.from.rou <- numeric(length=N)
    tries.from.rou <- numeric(length=N)
    a <- 0
    b <- sqrt(30) / 4
    c <- 0
    d <- 0.811
    for (i in 1:N) {
      sample <- rou.sample(a, b, c, d)
      samples.from.rou[i] <- sample[1]
      tries.from.rou[i] <- sample[2]
    }
    ```
    
    (v). **Probability of acceptance**
    
    The theoretical probability for acceptance is defined as $\frac{1}{2(b-a)(d-c)}$.
    
    ```{r}
    theoretical_p <- 1 / (2 * (b - a) * (d - c))
    empirical_p <- N / sum(tries.from.rou)
    sprintf("Theoretical probability %f", theoretical_p)
    sprintf("Empirical probability %f", empirical_p)
    ```
    
    (vi). **Using $U[0, 1]$ as the proposal distribution, generate 10000 samples from $f(x)$.**
    
    The proposal distribution $U[0, 1]$ has the same support as $f(x)$. To find
    the upper bound $c$ of the ratio of the target and the proposal distributions:
    
    $$ \sup_{x\in[0, 1]} \frac{f(x)}{g(x)} \leq c$$
    
    Since $g(x) = 1 \forall x \in [0, 1]$,
    
    $$\therefore c = \max_{x\in[0, 1]}f(x) = \max_{x \in [0, 1]}30x^2(1-x)^2 \approx 1.875$$
    
    ```{r}
    fx <- function(x) 30 * (x ^ 2) * ((1 - x)^2)
    
    uniform.sample <- function(paramC) {  # Sample from uniform for f(x) using AR
      n.try <- 0
      while (TRUE) {
        n.try <- n.try + 1
        proposal <- runif(1)
        
        # probability under proposal distro
        qy <- 1
        
        # probability under target distro
        py <- fx(proposal)
        
        if (runif(1) < py / (paramC * qy)) {
          break
        }
      }
      return(c(proposal, n.try))
    }
    paramC <- 1.875
    samples.from.uniform <- numeric(length=N)
    tries.from.uniform <- numeric(length=N)
    for (i in 1:N) {
      x <- uniform.sample(paramC)
      samples.from.uniform[i] <- x[1]
      tries.from.uniform[i] <- x[2]
    }
    ```
    
    (vii). **Density curve of target vs density curve of samples for both approaches.**
    
    For the Ratio of Uniforms method:
    
    ```{r}
    support <- seq(0, 1, length.out=N)
    plot(density(samples.from.rou), col="green",
         main="Ratio of Uniforms: Empirical vs Ideal Density")
    lines(support, fx(support), col="blue", lty=2)
    legend("topright", lty=1:2, col=c("green", "blue"), legend=c("Empirical", "Ideal"))
    ```
    
    For sampling from $U[0, 1]$ with AR method:
    
    ```{r}
    plot(density(samples.from.uniform), col="green",
         main="AR Sampling from U[0, 1]: Empirical vs Ideal Density")
    lines(support, fx(support), col="blue", lty=2)
    legend("topright", lty=1:2, col=c("green", "blue"), legend=c("Empirical", "Ideal"))
    ```
    
    (viii). **Comment on the efficiency of the above two methods.**
    
    ```{r}
    rou.prob.accept <- N / sum(tries.from.rou)
    ar.prob.accept <- N / sum(tries.from.uniform)
    sprintf("P(accept) for RoU: %f", rou.prob.accept)
    sprintf("P(accept) for AR from U[0, 1]: %f", ar.prob.accept)
    ```
    
    As seen, the AR method is more efficient, since it has a higher probability of acceptance.
    
---

2. Auditing co. wants to investigate false returns of taxpayers. $X$ denotes total
  number of false returns. Excessive funds follow $\mathcal{N}(\mu, \sigma^2)$.
  With probability $\delta$, taxpayers file zero false returns.
  
    (i) **Distribution of $X$ with parameters
    $\mu, \sigma$ and $\delta$, where $0 \leq \delta \leq 1$.**
    
    The mixture distribution is
    
    $$ f(x) = (1 - \delta) \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2}$$
    
    (ii). **Zero-inflated Normal Distribution**
    
    This is a 2 component mixture distribution such that,
    it is $~\mathcal{N}(\mu, \sigma^2)$ with probability $1 - \delta$ and
    zero with probability $\delta$.
    
    ```{r}
    ZIN.sample <- function(delta, mu, sigma) {
      x <- runif(1)
      if (x < delta) {
        sample <- 0
      } else {
        sample <- rnorm(1, mean=mu, sd=sigma)
      }
      return(sample)
    }
    ```
    
    (iii). **Generate 10000 samples with mu = 10, sigma = 1 and delta = 0.1.**
    ```{r}
    N <- 10000
    delta <- 0.1
    mu <- 10
    sigma <- 1
    samples.zin <- numeric(N)
    for (i in 1:N) {
      samples.zin[i] <- ZIN.sample(delta, mu, sigma)
    }
    ```
    
    (iv). **Histogrtam of the above samples**
    ```{r}
    hist(samples.zin, main="Zero-Inflated Normal Distribution")
    ```
    
    (v). **Compare empirical and theorectical means.**
    
    The theoretical mean is $\approx \mu (1 - \delta)$.
    
    ```{r}
    sprintf('Theoretical mean: %f', mu * (1 - delta))
    sprintf('Empirical mean: %f', mean(samples.zin))
    ```
    
    Both are approximately equal.

---

3. Generating samples from an $F(v_1, v_2)$ distribution using the known
relationship between $F$ and a $\chi^2$ distribution.

    (i). **Function to generate a $\chi^2_k$.**
    
    The sum of the squares of $k$ samples drawn from a standard normal
    ($\mathcal{N}(0, 1)$) distribution simulates a draw from a $\chi^2_{k}$ distribution.
    
    ```{r}
    chi.sq <- function(k) {
      normals <- rnorm(k)
      return(sum(normals ^ 2))
    }
    ```
    
    (ii). **Function to generate $F(v_1, v_2)$.**
    
    Suppose $X$ is a random sample from an $F$ distribution, $X \sim F(v_1, v_2)$. Then, $X$ can be approximated as 
    
    $$X = \frac{Y_1 / v_1}{Y_2/ v_2} $$,
    
    where $Y_1$ and $Y_2$ are _independent_ draws from two $\chi^2_{k}$
    distributions, with degrees of freedom $k = v_1$ and $k = v_2$, i.e.
    
    $$ Y_1 \sim \chi^2_{v_{1}}$$
    $$ Y_2 \sim \chi^2_{v_{2}}$$
    
    We can use this relationship as follows:
    
    
    ```{r}
    F.dist <- function(v1, v2) {
      y1 <- chi.sq(v1)
      y2 <- chi.sq(v2)
      return((y1 / v1) / (y2 / v2))
    }
    ```
    
    (iii). **Draw sample of size 10000 from $F(5, 7)$**
    
    ```{r}
    N <- 10000
    F.samples <- numeric(N)
    v1 <- 5
    v2 <- 7
    for (i in 1:N) {
      F.samples[i] <- F.dist(v1, v2)
    }
    ```
    
    (iv). **Compare empirical and theoretical means.**
    
    The theoretical mean of an $F(v_1, v_2)$ distribution is $\frac{v_2}{v_2 - 2}$
    
    ```{r}
    sprintf('Theoretical mean: %f', v2 / (v2 - 2))
    sprintf('Empirical mean: %f', mean(F.samples))
    ```
    
    Both are close to each other.
    
    (v). **Ideal vs Empirical Density.**
    
    ```{r}
    support <- seq(0, 10, length.out=N)
    plot(density(F.samples), xlim=c(0, 10), col="green",
         main="F distribution: Empirical vs Ideal Density")
    lines(support, df(support, v1, v2), col="blue", lty=2)
    legend("topright", lty=1:2, col=c("green", "blue"), legend=c("Empirical", "Ideal"))
    ```

---

